---
title: "Assignment 4 — DATA301"
author: "P. Javier Lopez Dondarza"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    code_folding: show
    self_contained: true
---

```{r global-options, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r setup}
library(fpp2)
library(ggplot2)
library(gridExtra)
set.seed(301)
```

# Exercise 1

## 1.1 Sample quantiles

  Wicklin (2017) compares the nine ways of computing a sample quantile that Hyndman & Fan (1996) list. They all interpolate between two neighbouring order statistics, and the only thing that changes is where the interpolation point sits. Types 1 to 3 are step functions that only return values from the sample. Types 4 to 9 are continuous and linear between order statistics. Type 6 is the default in Minitab and SPSS, type 2 in SAS, and type 7 in R. For large $n$ they all give about the same answer, but for small $n$ (and for $p$ close to 0 or 1) they can differ quite a bit.

I would use type 8. Hyndman & Fan recommend it because it is approximately median-unbiased whatever the distribution is, and that matters here: the exponential is skewed and I am going to use $n=5$. It is also continuous, so the IQR changes smoothly with the data. As a check I also run type 7, which is what R uses by default (`quantile(x, type = 7)`). NumPy's default (`method="linear"`) is type 7 as well.

```{r}
quantile(c(2, 4, 4, 5, 9), c(.25, .75), type = 7)
quantile(c(2, 4, 4, 5, 9), c(.25, .75), type = 8)
```

## 1.2 Estimators and bootstrap bias correction

For an exponential variable the quantile function is $q_p=-\ln(1-p)/\lambda$, so
$$q_{3/4}-q_{1/4}=\frac{\ln 4-\ln(4/3)}{\lambda}=\frac{\ln 3}{\lambda},$$
which is where $\hat\lambda_{IQR}=\ln 3/\text{IQR}$ comes from.

If $\hat\lambda_{IQR}$ is biased, then $b=\mathbb{E}[\hat\lambda_{IQR}]-\lambda$. The bootstrap idea is to replace the unknown distribution by the empirical one. I take $B$ resamples $X^{*b}$ of size $n$ with replacement, compute $\hat\lambda^{*b}_{IQR}$ for each, and estimate the bias by
$$\hat b_{boot}=\frac1B\sum_{b=1}^B\hat\lambda^{*b}_{IQR}-\hat\lambda_{IQR}(X).$$
Subtracting it gives the corrected estimator
$$\tilde\lambda=\hat\lambda_{IQR}-\hat b_{boot}=2\hat\lambda_{IQR}(X)-\frac1B\sum_{b=1}^B\hat\lambda^{*b}_{IQR},\qquad B=200.$$

One practical problem: with $n=5$ a resample can have a zero IQR (for example when the same value is drawn several times), which makes $\hat\lambda^*$ infinite. I drop the non-finite replicates before averaging.

```{r estimators}
QTYPE <- 8

lam_ml  <- function(x) 1 / mean(x)

lam_iqr <- function(x, type = QTYPE) {
  q <- quantile(x, c(.25, .75), type = type, names = FALSE)
  log(3) / (q[2] - q[1])
}

lam_bc <- function(x, B = 200, type = QTYPE) {
  n <- length(x)
  est <- lam_iqr(x, type)
  boot <- replicate(B, lam_iqr(sample(x, n, replace = TRUE), type))
  boot <- boot[is.finite(boot)]
  2 * est - mean(boot)
}

x <- rexp(30, rate = 1)
c(ML = lam_ml(x), IQR = lam_iqr(x), BC = lam_bc(x))
```

## 1.3 Monte Carlo experiment

The metric is the mean square error, $\text{MSE}(\hat\theta)=\mathbb{E}[(\hat\theta-\lambda)^2]$, estimated by $\frac1R\sum_{r=1}^R(\hat\theta_r-\lambda)^2$.

**How I chose $R$.** The Monte Carlo standard error of that estimate is $\text{sd}(e^2)/\sqrt R$ with $e=\hat\theta-\lambda$ (Robert & Casella, 2009). I fixed a target: the standard error should be about 2% of the MSE itself. Solving for $R$ gives $R=(\text{sd}(e^2)/(0.02\,\text{MSE}))^2$, and I estimate $\text{sd}(e^2)$ and MSE with a pilot run of 500 replications. Two observations shape the design:

* All three estimators are scale equivariant, $\hat\lambda(cX)=\hat\lambda(X)/c$. That means $\text{MSE}/\lambda^2$ is the same for every $\lambda$, so the $R$ I need depends on $n$ only. The three values of $\lambda$ then work as a check: the results should agree up to simulation noise.
* Small $n$ needs more replications. Both $1/\bar X$ and $\hat\lambda_{IQR}$ have heavy right tails, so the squared errors are very spread out.

For each $n$ I take the largest $R$ required by the three estimators, so they can all use the same samples.

```{r montecarlo}
one_cell <- function(lambda, n, R, B = 200) {
  res <- matrix(NA_real_, R, 3, dimnames = list(NULL, c("ML", "IQR", "BC")))
  for (r in 1:R) {
    x <- rexp(n, lambda)
    res[r, ] <- c(lam_ml(x), lam_iqr(x), lam_bc(x, B))
  }
  res
}

pilot_R <- function(n, R0 = 500, prec = 0.02, cap = 20000) {
  e <- (one_cell(1, n, R0) - 1)^2
  need <- (apply(e, 2, sd) / (prec * colMeans(e)))^2
  min(cap, ceiling(max(need) / 500) * 500)
}

Rn <- sapply(c(5, 30, 100), pilot_R)
names(Rn) <- c(5, 30, 100)
Rn
```

```{r montecarlo-run}
grid <- expand.grid(lambda = c(.1, 1, 10), n = c(5, 30, 100))

results <- do.call(rbind, lapply(seq_len(nrow(grid)), function(i) {
  lam <- grid$lambda[i]; n <- grid$n[i]; R <- Rn[as.character(n)]
  est <- one_cell(lam, n, R)
  e2 <- (est - lam)^2
  data.frame(lambda = lam, n = n, R = R,
             estimator = colnames(est),
             MSE = colMeans(e2),
             MC_se = apply(e2, 2, sd) / sqrt(R),
             relMSE = colMeans(e2) / lam^2)
}))

wide <- reshape(results[, c("lambda", "n", "R", "estimator", "MSE")],
                idvar = c("lambda", "n", "R"), timevar = "estimator",
                direction = "wide")
knitr::kable(wide, digits = 4)
```

```{r montecarlo-plot}
ggplot(results, aes(factor(n), relMSE, fill = estimator)) +
  geom_col(position = "dodge") +
  facet_wrap(~ lambda, labeller = label_both) +
  scale_y_log10() +
  labs(x = "n", y = "MSE / lambda^2 (log scale)")
```

**Results.** ML gana en las nueve celdas, sin excepción. La IQR es consistentemente peor que ML, y la corrección bootstrap (BC) es la peor de las tres en todos los casos, y con n = 5 es catastrófica: MSE de 263 a más de 121 000, frente a 0.006–62 de ML. Eso confirma dos cosas de la teoría: ML es el más eficiente cuando el modelo es correcto (usa toda la muestra), y la IQR pierde eficiencia porque solo usa dos cuantiles. La corrección de sesgo, en cambio, no ayuda aquí: con n pequeño, el IQR muestral puede acercarse a cero en algunos remuestreos bootstrap, y como $\tilde\lambda = 2\hat\lambda_{IQR}-\overline{\lambda^*}$ duplica el término original, cualquier inestabilidad en $\hat\lambda_{IQR}(X)$ se amplifica al cuadrado. Con n = 30 y n = 100 el problema es mucho más leve, pero BC sigue sin mejorar sobre la IQR sin corregir: reduce el sesgo a costa de más varianza, y aquí pierde ese cambio.

El ranking no cambia con λ, que es justo lo que predice la equivarianza de escala: dividiendo el MSE por λ² (columna relMSE), los tres λ dan valores parecidos dentro de cada n (hay algo de ruido de Monte Carlo, más visible en n = 5 donde R topó con el cap). 

For reference, $\mathbb{E}[\hat\lambda_{ML}]=\frac{n}{n-1}\lambda$ and $\text{MSE}(\hat\lambda_{ML})=\lambda^2\frac{n+2}{(n-1)(n-2)}$ for $n>2$, so the simulated MSE of the ML estimator should be close to that.

# Exercise 2

```{r ausbeer}
frequency(ausbeer)
start(ausbeer); end(ausbeer)
```

**1. Frequency.** It is 4, one observation per quarter (`frequency(ausbeer)` returns 4). This will also be visible in the time plot below, as a repeating up and down movement that completes once a year, so four points per cycle.

```{r tsplot}
autoplot(ausbeer) +
  labs(y = "Megalitres", title = "Quarterly beer production, Australia")
```

**2. Overall behaviour and trend.** There is a clear long-term trend, but it is not linear. Production rises fast until the mid-1970s, then it stays roughly flat until the early 1990s, and after that it drifts down towards 2010.

**4. Variability.** It is not constant. The seasonal swings get bigger when the level of the series is higher, which is typical of a multiplicative pattern. A log or Box–Cox transformation should fix that.

```{r outliers}
d <- data.frame(t = as.numeric(time(ausbeer)), y = as.numeric(ausbeer),
                q = cycle(ausbeer))
d$ratio <- d$y / stats::filter(d$y, rep(1/4, 4), sides = 2)
d[order(abs(log(d$ratio)), decreasing = TRUE)[1:6], ]
```

**5. Unusual observations.** Las seis que más se apartan del promedio local son:

1988.75, 1990.75 y 1991.75 (todas Q4): entre un 22% y un 25% por encima del promedio de su entorno. Coinciden con el pico que se ve en el gráfico de series justo antes de que la producción empiece a bajar, a finales de los 80.
1957.75 y 1958.75 (Q4): un 20-21% por encima, al principio de la serie.
1956.25 (Q2): un 17% por debajo, la única de las seis que está baja en vez de alta.

```{r seasonal}
ggseasonplot(ausbeer, year.labels = FALSE) + labs(title = "Seasonal plot")
ggseasonplot(ausbeer, polar = TRUE)
```

**3. Seasonality.** Yes. The same yearly shape appears again and again: one quarter is always the peak (Q4, which is spring/summer in Australia) and Q2 (winter) is the lowest.

**6. Seasonal plot.** Every year is one line across the four quarters. It makes clear that the shape (dip in Q2, peak in Q4) is the same every year, and that what changes is the height of the whole line: it goes up over the years and later comes back down. In the ordinary time plot this is harder to see because the trend takes most of the space.

```{r subseries}
ggsubseriesplot(ausbeer) + labs(title = "Seasonal subseries plot")
```

**7. Subseries.** There is one panel per quarter, with a horizontal line at its mean. Q4 has the highest mean and Q2 the lowest. Inside each panel the series is not flat, since all quarters go up and then down together. So the seasonal pattern is stable in shape, but the size of the differences between quarters grows with the level.

```{r transform}
lam <- BoxCox.lambda(ausbeer); lam
p1 <- autoplot(ausbeer) + ggtitle("Original")
p2 <- autoplot(log(ausbeer)) + ggtitle("Log")
p3 <- autoplot(BoxCox(ausbeer, lam)) +
  ggtitle(sprintf("Box-Cox (lambda = %.2f)", lam))
grid.arrange(p1, p2, p3, ncol = 1)
```

```{r detrend}
grid.arrange(
  autoplot(diff(log(ausbeer), lag = 4)) + ggtitle("Seasonal difference of log (lag 4)"),
  autoplot(diff(diff(log(ausbeer), lag = 4))) + ggtitle("Seasonal + first difference"),
  ncol = 1)
ggAcf(ausbeer)
```

**8. Detrend or transform?** I would do both before modelling. The trend is strong and not linear, so the series is not stationary in the mean. The seasonal amplitude also grows with the level, and the log or Box–Cox plot above shows it becomes much more even after transforming. The differenced series then look much more stable, and the ACF of the untransformed series shows large, slowly-decaying spikes at the seasonal lags (4, 8, 12...), which is more evidence that both the trend and the seasonality need to be removed before fitting a model.

# References

Hyndman, R. J., & Athanasopoulos, G. (2018). *Forecasting: Principles and practice*. OTexts. OTexts.com/fpp2

Hyndman, R. J., & Fan, Y. (1996). Sample quantiles in statistical packages. *The American Statistician*, 50(4), 361. https://doi.org/10.2307/2684934

Robert, C. P., & Casella, G. (2009). *Introducing Monte Carlo methods with R*. Springer.

Wicklin, R. (2017). Sample quantiles: A comparison of 9 definitions. SAS Blogs. https://blogs.sas.com/content/iml/2017/05/24/definitions-sample-quantiles.html

**Use of AI.** I used Claude (Anthropic) to help draft the structure, the code and the first version of the written explanations. I ran the code, checked the results and edited the text myself.
